---
title: "q3 stat 208"
output: html_document
date: "`r Sys.Date()`"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(cluster)
library(factoextra)
library(dendextend)
```


```{r}
df_genres <- read.csv("~/Desktop/data_by_genres.csv")
df_artists <- read.csv("~/Desktop/data_by_artist.csv")
```



```{r}
# Remove non-numeric or irrelevant columns
genres_data <- df_genres %>%
  select(-mode, -genres, -key)

artists_data <- df_artists %>%
  select(-mode, -artists, -count, -key)

# Scale the data
genres_scaled <- scale(genres_data)
artists_scaled <- scale(artists_data)
```

```{r}
set.seed(123)
fviz_nbclust(genres_scaled, kmeans, method = "wss") + 
  ggtitle("Elbow Method for Optimal Clusters (Genres)")

k_genres <- kmeans(genres_scaled, centers = 4, nstart = 25)
fviz_cluster(k_genres, data = genres_scaled, geom = "point", main = "K-means Clustering (Genres)")
```
```{r}
set.seed(123)
sample_size <- 1000  # You can try smaller (500) or slightly larger if your RAM allows

# Take a sample before PCA
sampled_artists <- artists_scaled[sample(nrow(artists_scaled), sample_size), ]

# Now do PCA and clustering on sampled data
pca_result <- prcomp(sampled_artists, scale. = TRUE)
artists_pca <- pca_result$x[, 1:5]

fviz_nbclust(artists_pca, kmeans, method = "wss") +
  ggtitle("Elbow Method after (PCA)")

k_artists <- kmeans(artists_pca, centers = 4, nstart = 25)
fviz_cluster(k_artists, data = artists_pca, geom = "point", main = "K-means Clustering")
```
```{r}
# Compute distance and perform clustering
dist_genres <- dist(genres_scaled, method = "euclidean")
hc_genres <- hclust(dist_genres, method = "ward.D2")

# Assign cluster labels (e.g., 4 clusters)
df_genres$cluster <- as.factor(cutree(hc_genres, k = 4))
```

```{r}
df_genres %>%
  group_by(cluster) %>%
  slice_max(order_by = popularity, n = 5) %>%
  select(cluster, genres, popularity, energy, danceability, valence) %>%
  arrange(cluster)
```

```{r}
# Compute mean of traits per cluster
cluster_means <- df_genres %>%
  group_by(cluster) %>%
  summarise(across(c(acousticness, danceability, energy, valence, popularity), mean, .names = "mean_{.col}"))

# Reshape using pivot_longer
cluster_means_long <- cluster_means %>%
  pivot_longer(cols = -cluster, names_to = "trait", values_to = "value")

# Plot
ggplot(cluster_means_long, aes(x = trait, y = value, fill = cluster)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Cluster Audio Trait Profiles", y = "Average Value", x = "Trait") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
set.seed(123)
sample_size <- 1000
sample_indices <- sample(nrow(artists_scaled), sample_size)
sampled_artists_scaled <- artists_scaled[sample_indices, ]
sampled_artists_raw <- df_artists[sample_indices, ]

pca_result <- prcomp(sampled_artists_scaled, scale. = TRUE)
artists_pca <- pca_result$x[, 1:5]  # First 5 principal components


dist_pca <- dist(artists_pca, method = "euclidean")
hc_pca <- hclust(dist_pca, method = "ward.D2")


artist_clusters <- cutree(hc_pca, k = 4)


sampled_artists_raw$cluster <- as.factor(artist_clusters)


cat("🔝 Top 5 Artists by Popularity in Each Cluster:\n")
sampled_artists_raw %>%
  group_by(cluster) %>%
  slice_max(order_by = popularity, n = 5) %>%
  select(cluster, artists, popularity, energy, danceability, valence) %>%
  arrange(cluster) %>%
  print(n = Inf)


# Average traits per cluster
cluster_means <- sampled_artists_raw %>%
  group_by(cluster) %>%
  summarise(across(c(acousticness, danceability, energy, valence, popularity), mean))

# Reshape for plotting
cluster_means_long <- cluster_means %>%
  pivot_longer(cols = -cluster, names_to = "trait", values_to = "value")

# Bar Plot
ggplot(cluster_means_long, aes(x = trait, y = value, fill = cluster)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Clustered Artist Profiles (Avg Audio Traits)", y = "Average Value", x = "Trait") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Interpretation of the Report:
1. Clustering Techniques Used

K-means clustering was applied to both genres and artists datasets.
Hierarchical clustering was also applied to the genre data, but only mentioned for artists in code/logic, not clearly shown or discussed in visual outputs.
artial coverage of the required methods (k-means = ✔, hierarchical = mostly genres).

🔹 2. Features Used for Clustering

Clustering was based on normalized audio traits:
acousticness, danceability, energy, valence, tempo, and most importantly, popularity.
orrect use of features aligning with the research goal.

🔹 3. Genre Clustering Outcomes

Genre data was clustered using both k-means and hierarchical clustering.
Visuals and tables show how genres grouped based on similarity in audio traits.
Popularity differences between clusters were analyzed (e.g., high-popularity clusters showed higher energy and danceability).
his directly supports the “finding high-performing groups” aspect of your question.

🔹 4. Artist Clustering Outcomes

Artist-level analysis used PCA + k-means and found distinct clusters.
Each cluster was summarized with average trait profiles, including popularity.
Representative (popular) artists were listed for each cluster.
his provides clear insight into which groups of artists are more successful in terms of popularity and traits.

🔹 5. Interpretability

Instead of messy dendrograms, the report smartly shows:
Top 5 genres or artists per cluster (by popularity)
Bar plots of cluster-level trait profiles
excellent choice for interpretability — results are understandable and actionable.